Adaptive broadcast tree for multicores

### 2013-05-07 paper.tex
Did try other topologies now
Scheduling as a separate section
Restructuring and general improvements

% New ideas
Instead of manually tuning algorithms to characteristics of individual
machines, we automatically configure our atomic broadcast
implementation based on a machine model.
% How will you go about it?
We will base our atomic broadcast on overlay networks, which we
automatically configure based on the machine model. One questions is
which information should actually be encoded in the machine model. %
Our idea is to then explore different approaches (such as MST
algorithms, clustering) to \emph{find good overlay networks} based on
the network model to base the atomic broadcast on. Then, for every
node, we need to decide which message to send first. This is a
\emph{scheduling} problem.

% How do we show it works
We auto-configure a wide range of different multicore machines and
show the performance of our implementation comes close to the
performance of hand-tuned implementations for all of these machines.

% Hypothesis
My hypothesis is that automatic tuning of atomic broadcast (and other
distributed algorithms) to machine characteristics is worth
doing. Obviously, we cannot achieve performance of hand-tuned
implementations, but we can get close enough to argue that loosing a
bit of performance to avoid tedious manual reconfiguration is worth if
for many applications (while others, where performance \emph{really}
matters, might still want to manually configure their stuff).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The multicore network model}

% selecting hard
Characteristics of the network influence the performance of
distributed algorithms. Due to the heterogeneous combinations of
complex networks on multicores, selection of algorithms is
particularly hard.

% model
We express the network characteristics in a multicore network
model. This model at this stage is by no means complete. We are going to
extend it over time to address more complexities of current and future
multicores.

% graph
We represent the multicore network as a graph. The nodes in this graph
correspond to cores, and edges the communication channels between
nodes. The machines we consider for now only have bi-directional
communication links. Other machines (like the SCC) have
uni-directional links, which is why we model the graph as directed
graph.

% shared memory -> fully meshed
This graph is fully meshed for shared memory machines. Every core can
communicate with every other node via the cache coherence protocol.

@@

@@

\begin{description}
\item[NUMA affinity] A list of cores on the same NUMA node
\item[synchrony] Parts of a multicore machine are synchronous. Hence,
  one characteristic of a node is an identifier for the synchrony
  cluster of a node.
\item[failure] Can nodes fail? If so, how? (Byzantine, crash)
\end{description}

These are currently not encoded in the model. The send and receive
cost are in part node specific. The faster a core is, the faster it
can to the multiplexing and marshalling. However, they are also link
dependent, so we view them as link characteristics.

@@

The cost of sending a message in a multicore machines does not so much
depend on the distance on the interconnect networks as on the number
of hops on such a path. Propagation on the wire is essentially for
free due to short distances between cores.

On a multi-hop path, however, packages traversing the interconnect
need to be buffered. This is expensive. Depending on the machine,
some routing is executed on every intermediate route, which increases
the latency further.

\begin{description}
\item[hop count] Number of hops to every other NUMA node.
\item[hop count histogram] For every NUMA node, we want to know how
  many other NUMA nodes can be reached with $n$ hops. This tells us
  something about how well such nodes are connected to other NUMA
  nodes. The intuition is that it makes sense to send a message to a
  well connected NUMA node early in time, to ``spread'' out a message
  as quickly as possible.
\end{description}

@@

\begin{description}
\item[breakdown] (quantitative) of message send cost:
  dominating propagation time vs.\ dominating send and receive time on
  multicores (per-node). Our model is currently limited to only the
  breakdown of the cost for sending messages. Maybe we do not strictly
  need to know about propagation time, as it is negligible independent
  of were messages are sent. Relevant seems to be the number of hops
  rather than the latency expressed in a concrete time.
\item[bounded delivery]
\item[reliability] (loss, link failure, in-order delivery, phantoms,
  corruption, duplicated messages): 
  multicore interconnect networks often are
\end{description}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Hierarchical model}
\label{sec:model_hierarchy}

Multicore machines are hierarchical. Processor cores are grouped in
NUMA nodes with shared caches. These NUMA nodes are then connected to
each other by increasingly complex interconnect networks.

We encode NUMA node affinity in our model. This allows to break down
programming of a multicore machine into two distinct instances. We
solve one instance of the problem between NUMA nodes, and then, in a
second step, within each NUMA node. These two steps can be executed
independently and different algorithms can be used in both case.

If cores on the same NUMA node share a cache, it makes sense to
leverage this hardware feature to implement shared-memory algorithms
on top of it. %
Communication across NUMA nodes is realized sending messages across
the packet-based interconnect network. Using message-passing based
implementations is the natural choice. %
Shared memory algorithms and message-passing based algorithms can be
composed together to program hierarchical machines similarly to what
was previously shown for composing of shared-memory
algorithms~\cite{Alistarh2012}.

@@

Each of the eight nodes consists of four cores. We approximate the
cost of sending messages within nodes as $1$ and across nodes as
$10*num(hops)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Atomic Broadcast for Multicores}

@@

\begin{figure}
\begin{tikzpicture}[>=latex,line join=bevel,scale=.5]
  \pgfsetlinewidth{.5bp}
\input{figs/final_gruyere_hierarchical}
\end{tikzpicture}
\caption{Hierarchical tree for a 8x4x1 multicore}
\label{fig:gruyere_hierarchy}
\end{figure}

@@

We discussed earlier in Section~\ref{sec:model_hierarchy} that
multicore machines are hierarchical.%

@@

Next, we evaluate ring topologies. Rings have the advantage that the
implicitly acknowledge messages once they arrive back to the
sender. This is beneficial for some algorithms such as quorums, where
nodes report back to the sender.

The downside is that some state is required on every node to remember
whether or not a message was received before. This however only works
if forwarding along the ring is only done after all local nodes
acknowledge a message. If messages are just forwarded as they arrive,
message transport does never terminate.

To add more parallelism, we add a second ring, which we send the
message on in parallel. The overlay ring network is visualized in
Figure~\ref{fig:gruyere_ring}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scheduling}

@@

Current numbers are listed in Table~\ref{tab:bc_measurements}.

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\item Try other topologies (Frans Kaashoek), e.g.\ rings
\item \emph{Load balancing in a tree} We could build several rings,
  that do not always include all coordinators and work with read- and
  write-sets. We can select these rings such that they nicely
  integrate with the physical topology. They can also have different
  sizes (e.g.\ a smaller one for read operations and larger ones for
  writes).
\item \emph{Aggregation for convergecast} %
  Convergecast is a way of collecting information from nodes (i.e.\ it
  is kind of the reverse of a broadcast). On the return path,
  information can be aggregated. One example is batching, where several
  pieces of information are grouped in the same message to reduce packet
  processing overhead. Another example is aggregation, i.e.\
  pre-processing of data. One example is pre-calculating the average
  value of child messages if the sink node is not interested in
  individual values. %
  Similarly to what has been done in wireless sensor networks (where it
  also matters to reduce the number of messages, but for other reasons:
  power consumption), we can do aggregation in nodes. In difference to
  traditional distributed systems, this works, because it is easy to
  deploy custom software on every node in the network. Classical
  distributed systems do not typically allow this. Furthermore, reducing
  the number of messages at the price of higher complexity does not make
  sense in classical systems. %
  Examples for aggregation: number of nodes agreeing to something, find
  capabilities (concatenate core ids). 
\end{itemize}

>>>>

% New ideas
Instead of manually tuning algorithms to characteristics of individual
machines, we automatically configure our atomic broadcast
implementation based on a machine model. We represent this model as an
annotated graph. One question we are looking at is which information
should actually be part of this machine model.

% How will you go about it?
We will base our atomic broadcast on \emph{overlay networks}, which we
automatically configure. We execute our configuration stage on a model
representing the concrete machine the algorithm is intended to run
on. %
Our idea is to then explore different approaches for finding overlay
networks based on the network model graph. Approaches we are currently
investigating are: minimum spanning tree algorithms, binary trees,
clustering techniques as well as treating the problem
hierarchically. Assuming the overlay network, we then need to decide
on every node which message to send first. This is a \emph{scheduling}
problem.

% How do we show it works
We auto-configure a wide range of different multicore machines and
compare the performance of our implementation with hand-tuned
implementations on all of these machines.
% Hypothesis
Our hypothesis is that automatic tuning of atomic broadcast (and other
distributed algorithms) to machine characteristics is worth
doing. Obviously, we cannot achieve performance of hand-tuned
implementations, but we hope to get close enough to argue that loosing
a bit of performance to avoid tedious manual reconfiguration is worth
if for many applications (while others, where performance
\emph{really} matters, might still want to manually configure their
stuff).

% three-step approach
We break the problem down into three steps:
\begin{inparaenum}[\itshape 1\upshape)]
\item We first represent multicore machines by a model. We describe
  this model in Section~\ref{sec:model},
\item We then find an overlay for a concrete network model. This
  overlay network describes a topology (such as a tree or a ring) for
  communication within the machine. Algorithms to select overlay
  networks are described in Section~\ref{sec:overlay}, and
\item Given the overlay, we need to decide for every node in which
  order to send messages. We investigate this scheduling problem in
  Section~\ref{sec:scheduling}.
\end{inparaenum}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The multicore network model}
\label{sec:model}

% selecting hard
Characteristics of the network influence the performance of
distributed algorithms. 
% model
We express these network characteristics in a multicore network
model. This model at this stage is by no means complete. We are going to
extend it over time to address more complexities of current and future
multicores.
% static
Our network model is \emph{static}, which in particular implies that
we do not consider run-time effect such as congestion for the moment.

% graph
We represent the multicore network as a \emph{directed graph}. The nodes in
this graph correspond to cores, and edges to communication channels
between nodes. The machines we consider for now only have
bi-directional communication links. Other machines (like the SCC) have
uni-directional links, which is why we model the graph as directed
graph.
% shared memory -> fully meshed
This graph is \emph{fully meshed} for shared memory machines. Every core can
communicate with every other node via the cache coherence protocol.

% machines hierarchical
Multicore machines are hierarchical. Processor cores are grouped in
NUMA nodes with shared caches. These NUMA nodes are then connected to
each other by increasingly complex interconnect networks. 
% NUMA affinity -> model
We encode NUMA node affinity in our model. This allows to break down
programming of a multicore machine into two distinct instances. We
solve one instance of the problem between NUMA nodes, and then, in a
second step, within each NUMA node. These two steps can be executed
independently and different algorithms can be used in both case.

% why it makes sense?
If cores on the same NUMA node share a cache, it makes sense to
leverage this hardware feature to implement shared-memory algorithms
on top of it. %
Communication across NUMA nodes is realized sending messages across
the packet-based interconnect network. Using message-passing based
implementations is the natural choice. %
Shared memory algorithms and message-passing based algorithms can be
composed together to program hierarchical machines similarly to what
was previously shown for composing of shared-memory
algorithms~\cite{Alistarh2012}.

@@

\stefan{Not sure yet if we really need this, but it should be helpful
  for treating the machine hierarchically}

@@

\begin{description}
\item[NUMA affinity] A list of cores on the same NUMA node.
\item[synchrony] Parts of a multicore machine are synchronous, i.e.\
  cores execute in lock-steps, where every node can send a message to
  every other node in one such step. This implies that there is a
  global time and message delivery time is bound. Both can be assumed
  on today's multicores. The global time comes from the fact that
  cores are driven by the same quartz or, if not, they are
  synchronized such that the jitter is less than a cycle, which leads
  to cores seeing the same time. %
  For these reasons, synchrony is not part of our current machine
  model. Within today's multicores, synchrony can be
  assumed. \stefan{Can it?}
\item[failure] Can nodes fail? If so, how? (Byzantine, crash). We do
  not consider failures at the moment, but believe that we will
  inevitably have to do so for future multicores, where it is unlikely
  that hardware can hide failures from software.
\end{description}

@@

\stefan{Maybe characteristics of NUMA nodes do not have to be encoded
  separately, but as part of the node characteristics?}

Some characteristics of multicore communication stem from the NUMA
hierarchy. Thus, such characteristics can be encoded separately in the
model.

\begin{description}
\item[hop count] Number of hops to every other NUMA node. The cost of
  sending a message in a multicore machines does not so much depend on
  the distance on the interconnect networks as on the number of hops
  on such a path. Propagation on the wire is essentially for free due
  to short distances between cores. %
  On a multi-hop path, however, packages traversing the interconnect
  need to be buffered. This is expensive. Depending on the machine,
  some routing is executed on every intermediate route, which
  increases the latency further.
\item[hop count histogram] For every NUMA node, we want to know how
  many other NUMA nodes can be reached with $n$ hops. This tells us
  something about how well such nodes are connected to other NUMA
  nodes. The intuition is that it makes sense to send a message to a
  well connected NUMA node early in time, to ``spread'' out a message
  as quickly as possible.
\end{description}

@@

\begin{description}
\item[cost breakdown] (quantitative) of message send cost:
  dominating propagation time vs.\ dominating send and receive time on
  multicores (per-node). Our model is currently limited to only the
  breakdown of the cost for sending messages. Maybe we do not strictly
  need to know about propagation time, as it is negligible independent
  of were messages are sent. Relevant seems to be the number of hops
  rather than the latency expressed in a concrete time.
\item[bounded delivery]
\item[reliability] (loss, link failure, in-order delivery, phantoms,
  corruption, duplicated messages): 
  multicore interconnect networks often are
\end{description}

@@

@@

Each of the eight NUMA nodes consists of four cores. We approximate the
cost of sending messages within nodes as $1$ and across nodes as
$10*num(hops)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overlay selection}
\label{sec:overlay}

@@

\begin{figure}
\begin{tikzpicture}[>=latex,line join=bevel,scale=.5]
  \pgfsetlinewidth{.5bp}
\input{figs/final_gruyere_hierarchical}
\end{tikzpicture}
\caption{Hierarchical tree for a 8x4x1 multicore. Nodes 0, 4, 8, 12,
  16, 20, 24 and 28 are coordinators. They are arranged in a binary
  tree starting at node 0. Communication within NUMA nodes is
  sequential (as an example: coordinator node 24 sends a message to
  all cores on the same NUMA node, namely 25, 26 and 27)}
\label{fig:gruyere_hierarchy}
\end{figure}

@@

\stefan{Have not implemented this, but it might be interesting to try
  out later}

We discussed earlier in Section~\ref{sec:model} that
multicore machines are hierarchical.%

@@

Next, we evaluate ring topologies. Rings have the advantage that the
implicitly acknowledge messages once they arrive back to the
sender. This is beneficial for some algorithms such as quorums, where
nodes report back to the sender. This, however, only works if forwarding
along the ring is only done after all local nodes acknowledge a
message.

The downside is that some state is required on every node to remember
whether or not a message was received before. If messages are just
forwarded as they arrive, message transport does never terminate.

To add more parallelism, we add a second ring, which we send the
message on in parallel. The overlay ring network is visualized in
Figure~\ref{fig:gruyere_ring}. Node that the link from node 20 to 8 is
a two hop link and, thus, has a cost of 20 instead of the usual 10.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Scheduling}
\label{sec:scheduling}

@@

\stefan{These numbers are outdated. Keeping them for reference}

Some numbers are listed in Table~\ref{tab:bc_measurements}.

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\item \emph{Load balancing in a tree} We could build several rings,
  that do not always include all coordinators and work with read- and
  write-sets. We can select these rings such that they nicely
  integrate with the physical topology. They can also have different
  sizes (e.g.\ a smaller one for read operations and larger ones for
  writes).
\item \emph{Aggregation for convergecast} %
  Convergecast is a way of collecting information from nodes (i.e.\ it
  is kind of the reverse of a broadcast). On the return path,
  information can be aggregated. One example is batching, where several
  pieces of information are grouped in the same message to reduce packet
  processing overhead. Another example is aggregation, i.e.\
  pre-processing of data. One example is pre-calculating the average
  value of child messages if the sink node is not interested in
  individual values. %
  Similarly to what has been done in wireless sensor networks (where it
  also matters to reduce the number of messages, but for other reasons:
  power consumption), we can do aggregation in nodes. In difference to
  traditional distributed systems, this works, because it is easy to
  deploy custom software on every node in the network. Classical
  distributed systems do not typically allow this. Furthermore, reducing
  the number of messages at the price of higher complexity does not make
  sense in classical systems. %
  Examples for aggregation: number of nodes agreeing to something, find
  capabilities (concatenate core ids). 
\end{itemize}
